---
title: "Public perception of climate change"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{climate change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


# Public perception of climate change

This short document analyses the climate change dataset originally
presented in the hyperdirichlet R package^[RKS Hankin 2010.  "A
generalization of the Dirichlet distribution", *Journal of Statistical
Software*, 33:11] but using the hyper2 package instead.


```{r}
library("hyper2",quietly=TRUE)
```

## Create a likelihood function

First specify the matrix as appearing in hyperdirichlet, as originally
provided by Saffy:

```{r}
M <- matrix(c(
    5 , 3 , NA,  4, NA,   3,
    3 , NA,  5,  8, NA,   2,
    NA,  4,  9,  2, NA,   1,
    10,  3, NA,  3,  4,  NA,
    4 , NA,  5,  6,  3,  NA,
    NA,  4,  3,  1,  3,  NA,
    5 ,  1, NA, NA,  1,   2,
    5 , NA,  1, NA,  1,   1,
    NA,  9,  7, NA,  2,   0)
  , byrow=TRUE,ncol=6)
colnames(M) <- c("NB","L","PB","THC","OA","WAIS")
M
```

(M is called ```icons_matrix``` in the package).  Each row of ```M```
corresponds to a particular cue given to respondents.  The first row,
for example, means that a total of $5+3+4+3=15$ people were shown
icons NB, L, TCH, WAIS [column names of the the non-NA entries]; 5
people chose NB as most concerning, 3 chose L, and so on.  The dataset
is more fully described in the hyperdirichlet package.  To recreate
the ```icons``` dataset in the ```hyper2``` package, we use the
```saffy()``` function:

```{r}
icons <- saffy(M)
icons
```

At this point, the ```icons``` object as created above is
mathematically identical to ```icons``` object in the hyperdirichlet
package (and indeed the hyper2 package), but the terms might appear in
a different order.

## Analysis of the icons dataset

The first step is to find the maximum
likelihood estimate for the ```icons``` likelihood:

```{r}
mic <- maxp(icons)
mic
dotchart(mic,pch=16)
```

We also need the likelihood at an unconstrained maximum:

```{r}
L1 <- loglik(icons,indep(mic))
L1
```

Agreeing to 4 decimal places with the value given in the
hyperdirichlet package.

## Hypothesis 1: $p_1\geqslant 1/6$

Following the analysis in Hankin 2010, we first observe that NB [the
Norfolk Broads] is the largest-probability icon (as expected on
theoretical grounds by O'Neill).  This would suggest that $p_1$ is
large, in some sense.  Consider the hypothesis that $p_1\leqslant 1/6$
and to that end perform a constrained optimization, with (active)
constraint that $p_1\leqslant 1/6$.  This is most easily done by
imposing additional constraints to the ```maxp()``` function via the
```fcm``` and ```fcv``` arguments:


```{r}
small <- 1e-6  #  ensure start at an interior point
maxlike_constrained <- 
    maxp(icons, startp=indep(equalp(icons))-c(small,0,0,0,0),
         give=TRUE, fcm=c(-1,0,0,0,0),fcv= -1/6)
maxlike_constrained	 
```	

So the extra support is given by

```{r}
ES <- L1-maxlike_constrained$value 
ES
```

(compare 2.608181 from the hyperdirichlet package).  This exceeds
Edwards's two-units-of-support criterion); a pvalue would be

```{r}
pchisq(2*ES,df=1,lower.tail=FALSE)
```

both of which would indicate that we may reject that hypothesis that
$p_1\leqslant 1/6$ and thereby infer $p_1>\frac{1}{6}$.

## Hypothesis 2: $p_1\geqslant\max\left(p_2,\ldots,p_6\right)$

Another constrained likelihood maximization, although this one is not
possible with convex constraints.  We have to compare ```L1``` against
the maximum likelihood over the region defined by $p_1$ being greater
than *at least one* of $p_2,\ldots,p_6$.  The union of convex sets is
not necessarily convex (think two-way Venn diagrams).  As far as I can
see, the only way to do it is to perform a sequence of five
constrained optimizations: $p_1\leqslant p_2, p_1\leqslant p_3,
p_1\leqslant p_4, p_1\leqslant p_5$.  The fillup constraint would be
$p_1\leqslant p_6\longrightarrow 2p_1+p_2+\ldots p_5\leqslant 1$.  We
then choose the largest likelihood from the five.

```{r}
o <- function(Ul,Cl,startp,give=FALSE){
    small <- 1e-6  #  ensure start at an interior point
    if(missing(startp)){startp <- small*(1:5)+rep(0.1,5)}			
    out <- maxp(icons, startp=small*(1:5)+rep(0.1,5), give=TRUE, fcm=Ul,fcv=Cl)
    if(give){
        return(out)
    }else{
        return(out$value)
    }
}

p2max <- o(c(-1, 1, 0, 0, 0), 0)
p3max <- o(c(-1, 0, 1, 0, 0), 0)
p4max <- o(c(-1, 0, 0, 1, 0), 0)
p5max <- o(c(-1, 0, 0, 0, 1), 0)
p6max <- o(c(-2,-1,-1,-1,-1),-1)
```

(the final line is different because $p_6$ is the fillup value; the
hyperdirichlet package deals with this issue 

```{r}
likes <- c(p2max,p3max,p4max,p5max,p6max)
likes
ml <- max(likes) 
ml
```

So the extra likelihood is given by

```{r}
L1-ml
```

(the hyperdirichlet package gives 0.0853 here, the small difference
probably being due to the difficulties of optimizing over a nonconvex
region).  It's worth looking at the evaluate too:

```{r}
o(c(-1, 1, 0, 0, 0), 0,give=TRUE)$par
o(c(-1, 0, 1, 0, 0), 0,give=TRUE)$par
o(c(-1, 0, 0, 1, 0), 0,give=TRUE)$par
o(c(-1, 0, 0, 0, 1), 0,give=TRUE)$par
o(c(-2,-1,-1,-1,-1),-1,give=TRUE)$par
```

## Low frequency responses

The next hypothesis was testing the smallness of $p_5+p_6$, and we
suspected that $p_5+p_6 < \frac{1}{3}$.  This translates into
optimizing subject to $p_5+p_6\geqslant\frac{1}{3}$, and the required
constraint is $-p_1-p_2-p_3-p_4\geqslant-\frac{2}{3}$ (because
$p_5+p_6=1-p_1-p_2-p_3-p_4$).  Dead easy:

```{r} 
jj <- o(c(-1,-1,-1,-1,0) , -2/3, give=TRUE,start=indep((1:6)/21))$value
jj
```

then the extra support is 

```{r}
L1-jj
```

(compare 7.711396 in hyperdirichlet, not sure why the discrepancy is
so large).

## Final example

The final example was
$\max\left\{p_5,p_6\right\}\geqslant\min\left\{p_1,p_2,p_3,p_4\right\}$.
This means the optimization is constrained so that *at least one* of
$\left\{p_5,p_6\right\}$ exceeds *at least one* of
$\left\{p_1,p_2,p_3,p_4\right\}$.  So we have the union of the various
possibilities:

\[
\bigcup_{j\in\left\{5,6\right\}\atop k\in\left\{1,2,3,4\right\}}
\left\{\left(p_1,p_2,p_3,p_4,p_5,p_6\right)\left|\sum p_i=1,
p_j\geqslant p_k\right.\right\}
\]

and of course $p_6\geqslant p_2$, say, translates to
$-p_1-2p_2-p_3-p_4-p_5\geqslant -1$.

```{r}
start <- indep(c(small,small,small,small,0.5-2*small,0.5-2*small))
jj <- c(
   o(c(-1, 0, 0, 0, 1), 0,start=start),
   o(c( 0,-1, 0, 0, 1), 0,start=start),
   o(c( 0, 0,-1, 0, 1), 0,start=start),
   o(c( 0, 0, 0,-1, 1), 0,start=start),

   o(c(-2,-1,-1,-1,-1),-1,start=start),
   o(c(-1,-2,-1,-1,-1),-1,start=start),
   o(c(-1,-1,-2,-1,-1),-1,start=start),
   o(c(-1,-1,-1,-2,-1),-1,start=start)
   )
jj
max(jj)
```

So the extra support is

```{r}
L1-max(jj)
```
(compare hyperdirichlet which gives 3.16, not sure why the difference).
We should look at the maximum value:

```{r}
   o(c( 0, 0, 0,-1, 1), 0,give=TRUE,start=start)
```

So the evaluate is at the boundary, for $p_4=p_5$.  I have no
explanation for the discrepancy between this and that in the
hyperdirichlet package.