---
title: "Ladies' figure skating"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{figure_skating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Ladies' figure skating in the 2002 Winter Olympics

This script creates and analyses hyper2 object "skating" which is a
likelihood function for the strengths of the competitors in the
Ladies' Figure Skating at the 2002 Winter Olympics.  Dataframe
```skating_table``` is copied from Lock and Lock (see skating.Rd for
more details).

```{r}
library("hyper2",quietly=TRUE)
data("skating")
skating_table
```		

Object ```skating_table``` is structured so that each competitor is a
row, and each judge is a column.  Function ```order_likelihood()```
considers each row to be a race [or a judge], so we need to take a
transpose.  The following R idiom is plausible but incorrect:


```{r}
skating_incorrect <- order_likelihood(t(skating_table))
```

The above is incorrect because we want the ranks, and we have the
order.  Correct analysis follows.


```{r}
skating_rank <- apply(skating_table,2,order)
head(skating_rank)
```

Column 1 of ```skating_table``` and column 1 of ```skating_rank``` are
confusingly simular.  So consider column 2 instead, corresponding to
judge number 2.  If ```a=skating_table[,2]``` and
```b=skating_rank[,2]``` we have ```a=c(4,1,3,2, ...)``` and
```b=c(2,4,3,1, ...)```.  Thus ```a``` means that hughes was judged to
be fourth best, slutskaya was judged to be the best, kwan third best,
cohen second best, and so on.  Conversely, ```b``` means that the best
competitor was number two [slutskaya], the second best number was
number four [cohen], the third best was number three [kwan], the
fourth best number 1 [hughes], and so on.  Note that this means the
rows of ```skating_rank``` cannot be named.


```{r}
skating <- order_likelihood(t(skating_rank))
pnames(skating) <- rownames(skating_table)
head(skating)
```

Now some data visualization.  First the MLE for the strengths:


```{r,dpi=72}
dotchart(maxp(skating))
dotchart(log(maxp(skating)))
```


Now compare my rating with the official point-tallying method:

```{r,dpi=72}
par(pty='s')
lik <- order(maxp(skating),decreasing=TRUE)
plot(1:23,lik,asp=1,pch=16,xlab='official order',ylab='likelihood order')
text(1:23,lik,asp=1,pch=16,pos=c(rep(4,19),rep(2,4)),rownames(skating_table))
abline(0,1)
```



## Evidence for medal positions

Looking at the figure, it seems that the medallists (Kwan, Slutskya,
Hughes) were considerably higher in strength than the rest of the
field.  Here I will test the hypothesis that the medallists were in
fact the strongest three competitors.  Technically you need to
optimize over the union of the possibilities that one of the three
medallists did not come in the top three; but this is hard.  We will
do something much easier but numerically equivalent: optimize of the
union of outcomes where either Cohen or Suguri had higher strength
than any of the medallists.

```{r}
jj <-
    matrix(c(
        -1, 0, 0,   1,0,  # Hughes   < Cohen
         0,-1, 0,   1,0,  # Slutskya < Cohen
         0, 0,-1,   1,0,  # Kwan     < Cohen
        
        -1, 0, 0,   0,1,  # Hughes   < Suguri
         0,-1, 0,   0,1,  # Slutskya < Suguri
         0, 0,-1,   0,1), # Kwan     < Suguri
        byrow=TRUE,ncol=5)

problem_constraints <-  # fill with zeros for other competitors
    cbind(jj,matrix(0,nrow(jj),size(skating)-ncol(jj)-1))

standard_constraints <- rbind(
    diag(22), # p_i >=0
    -1        # p_1+p_2+...+p_22 <= 1
    )

small <-1e-3  # need a sensible start value satisfying the constraints
start <- c(rep(2*small,3),rep(3*small,2),rep(small,17))

objective <- function(p){loglik(skating,p)}
grad <- function(p){gradient(skating,p)}

out <- rep(0,nrow(problem_constraints))
fullout <- list()
for(i in seq_len(nrow(problem_constraints))){
    UI <-   # constraints are: ONE of the problem constraints,
            # and all of the standard constraints
      rbind(problem_constraints[i,,drop=FALSE],
      standard_constraints)
    CI <-
        c(
            0,                       # p_i >=0
            rep(0,size(skating)-1),  # p_1+p_2+...+p_22 <= 1
            -1                       # problem constraint
        )
    
    jj <-
        constrOptim(
            theta   = start,
            f       = objective,
            grad    = grad,
            ui      = UI,
            ci      = CI,
            control = list(fnscale = -1)  # maximizing
        )
    fullout[[i]] <- jj
    out[i] <- -jj$value   # minus because of fnscale
}

max(out)-maxp(skating,give=TRUE)$value 
```

that is, almost 6 units of support, well above the two units suggested
by Edwards.