---
title: "Ladies' figure skating"
author: "Robin K. S. Hankin"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{figure_skating}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Ladies' figure skating in the 2002 Winter Olympics

This script creates and analyses hyper2 object "skating" which is a
likelihood function for the strengths of the competitors in the
Ladies' Figure Skating at the 2002 Winter Olympics.  Dataframe
```skating_table``` is copied from Lock and Lock (see skating.Rd for
more details).

```{r}
library("hyper2")
data("skating")
skating_table
```		

Object ```skating_table``` is structured so that each competitor is a
row, and each judge is a column.  Function ```order_likelihood()```
considers each row to be a race [or a judge], so we need to take a
transpose.  The following R idiom is plausible but incorrect:


```{r}
skating_incorrect <- order_likelihood(t(skating_table))
```

The above is incorrect because we want the ranks, and we have the
order.  Correct analysis follows.


```{r}
skating_rank <- apply(skating_table,2,order)
head(skating_rank)
```

Column 1 of ```skating_table``` and column 1 of ```skating_rank``` are
confusingly simular.  So consider column 2 instead, corresponding to
judge number 2.  If ```a=skating_table[,2]``` and
```b=skating_rank[,2]``` we have ```a=c(4,1,3,2, ...)``` and
```b=c(2,4,3,1, ...)```.  Thus ```a``` means that hughes was judged to
be fourth best, slutskaya was judged to be the best, kwan third best,
cohen second best, and so on.  Conversely, ```b``` means that the best
competitor was number two [slutskaya], the second best number was
number four [cohen], the third best was number three [kwan], the
fourth best number 1 [hughes], and so on.  Note that this means the
rows of ```skating_rank``` cannot be named.


```{r}
skating <- order_likelihood(t(skating_rank))
pnames(skating) <- rownames(skating_table)
head(skating)
```

Now some data visualization.  First the MLE for the strengths:


```{r}
dotchart(maxp(skating))
dotchart(log(maxp(skating)))
```


Now compare my rating with the official point-tallying method:

```{r}
par(pty='s')
lik <- order(maxp(skating),decreasing=TRUE)
plot(1:23,lik,asp=1,pch=16,xlab='official order',ylab='likelihood order')
text(1:23,lik,asp=1,pch=16,pos=c(rep(4,19),rep(2,4)),rownames(skating_table))
abline(0,1)
```

