---
title: "Repeated incomplete rank observations"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Suppose there are five people in a family.  Each day for five days,
one person cooks a meal.  After the fifth day, each person considers
the four meals cooked by other family members, and puts them in order
of preference.  This process is repeated and here I consider how to
interpret such data.

```{r loadlib}
library(hyper2)
H <- hyper2(pnames=paste("p",1:5,sep=""))
H
```

Now add some data:

```{r data}
H <- H +
	# 3 April 2020:
    rank_likelihood(c(5,4,3,2)) + # person 1 says 5 is the best, 2 the worst.
    rank_likelihood(c(2,5,1,4)) + # person 2 says 1 is the best, 4 the worst.
    rank_likelihood(c(5,1,2,3)) + # person 3 says 1 is the best, 5 the worst.
    rank_likelihood(c(3,5,4,1)) + # person 4 says 1 is the best, 5 the worst.
    rank_likelihood(c(3,2,1,4)) + # person 5 says 2 is the best, 4 the worst.

	# 10 April 2020:
    rank_likelihood(c(5,3,4,1)) + # person 1 says 5 is the best, 4 the worst.
    rank_likelihood(c(2,3,4,1)) + # person 2 says 1 is the best, 4 the worst.
    rank_likelihood(c(3,5,2,1)) + # person 3 says 1 is the best, 5 the worst.
    rank_likelihood(c(2,4,5,3)) + # person 4 says 1 is the best, 5 the worst.
    rank_likelihood(c(2,5,4,1)) + # person 5 says 2 is the best, 4 the worst.

H
```

Note that noone is allowed to vote for themselves.  There are
$5\cdot4!=120$ distinct observations.  We may find maximum likelihood
estimate for the strengths,
$\operatorname{argmax}\mathcal{L}\left(p_1,\ldots,p_5\right)$:


```{r}
(mH <- maxp(H))
pie(mH)
```


