---
title: "curling"
output: html_document
---

```{r setup, include=FALSE}
set.seed(0)
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
options("digits" = 5)
```

## Curling


Curling is a sport with stones and brooms.  Following dataset taken
from [wikipedia](https://www.google.com "Google's Homepage"), women's
tournament:

```{r readcurlingtable}
curling_table  <- read.table("curling.txt",header=TRUE)
curling_table
```

Thus we see that Canada came first in 1998, third in 2002, and so on;
in 1998, Canada came first, China did not compete [`NA` meaning did
not compete], Denmark came second, and so on.

First we will interpret `NA` in the same way as `DNS` in the context
of formula 1 motor racing.  We need to define a new data frame, `b`:

```{r,defineb}
ct <- curling_table
ct[is.na(ct)] <- "DNS"
ct
```

In this form we can easily convert to a support function:

```{r,definsupportb,cache=TRUE}
H1 <- ordertable2supp(ct)
H1
mH1 <- maxp(H1)
pie(mH1)
```

## Interpretation of `NA`

There is a different interpretation of `NA` entries.  We might hold
that no inference should be made from a team's entry being `NA` entry
because that that team did not compete.  This means that function
`ordertable2supp()` is not appropriate and we need to use a different
method:

```{r,makehyper2curling}
H2 <- hyper2(pnames=rownames(curling_table))

for(i in seq_len(ncol(curling_table))){
    x <- curling_table[,i]
    names(x) <- rownames(curling_table)
    H2 %<>% order_obs(x[!is.na(x)])
}
head(H2)
(mH2 <- maxp(H2))
pie(mH2)
```


##  Comparison

```{r,plotcomplog}
par(pty="s")
plot(log10(mH1),log10(mH2),asp=1,pch=16,xlim=c(-2.6,0),ylim=c(-2.6,0))
posn <- c(Canada=2, China=4, Denmark=4, Germany=4, GB=4, Italy=1, Japan=2, Norway=2,
          Russia=4, South_Korea=4, Sweden=4, Switzerland=2, US=4)
for(i in seq_len(nrow(curling_table))){
	text(log10(mH1)[i],log10(mH2)[i],rownames(curling_table)[i],col="gray",pos=posn[i])
}
abline(0,1)
grid()
```

The two likelihood systems have very different evaluates.
