---
title: "exponential Bradley-Terry: likelihood-based methods on a grid of values"
author: "robin hankin"
date: "2025-04-02"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
library("partitions")
library("magrittr")
set.seed(0)
```

<font size="1">
(takes maybe two or three hours to run without cache; each
<tt>parkrunlikefunobs</tt> chunk takes maybe thirty minutes).
</font>

First we define `BT()` which specifies Bradley-Terry strengths $\alpha x^i$:

```{r defineBT}
BT <- function(n,x){
    out <- x^(0:(n-1))
    if(x != 1){
        out <- out*(1-x)/(1-x^n)
    } else {
        out <- out/n
    }
    names(out) <- paste0("p",str_pad(1:n,ceiling(log10(n))))
    out
}
BT(4,0.5)
sum(BT(4,0.5))
```

We will consider $n=9$, $x=0.5$, $r=4$ and make repeated observations:


```{r showtabulate}
(jj <- BT(9,0.5))
o <- tabulate(replicate(1e4,which(rrace(jj) == "p4")))
o
plot(o,type='h')
```


    
```{r definegetprob, caasdf=TRUE}
getprob <- function(n, r, x, N=1e4){
    jj <- BT(n, x)
    wanted <- paste0("p", str_pad(r, ceiling(log10(n))))
    c(r=r, x=x, setNames(tabulate(replicate(N, which(rrace(jj) == wanted)),nbins=n), paste0("p", 1:n)))
}
```

```{r makeII, cache=TRUE}
r_try <- 1:9
x_try <- seq(from = 0.1, to = 1, by = 0.05)
jj <- as.matrix(expand.grid(r_try, x_try))
II <- t(apply(jj, 1, function(v){getprob(9, v[1], v[2], N=1e4)}))
head(II)
```

```{r showIItail}
tail(II,3)
```

Above we see object `II`.  We use this to calculate probabilities of
the focal competitor placing $i$, conditional on $r$ and $x$.  It
might have been better to divide columns `pi` through `p9` by `1e4` to
get probabilities.  To see contours, look at `exponential_BT.Rmd`.


```{r makesupp, cache=TRUE}
make_obs <- function(r_true=3, x_true=0.85, nobs=100){

    set.seed(0)
    o <- getprob(9, r_true, x_true, obs)  # observations
    return(o)
}


support_func <- function(o, v){
    v <- v[-(1:2)]
    v <- v/sum(v)
    dmultinom(o[-(1:2)], prob=v, log=TRUE)
}
support <- rep(0,nrow(II))
for(i in seq_len(nrow(II))){
    support[i] <- support_func(o, II[i,])
}
```


Now

```{r}
X <- cbind(support,II)
head(X)
tail(X)
```


```{r}
m <- which.max(support)
X[(m-5):(m+5),]
```
