---
title: "exponential Bradley-Terry: likelihood-based methods on a grid of values"
author: "robin hankin"
date: "2025-04-02"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
library("partitions")
library("magrittr")
```

<font size="1">
(takes maybe two or three hours to run without cache; each
<tt>parkrunlikefunobs</tt> chunk takes maybe thirty minutes).
</font>

First we define `BT()` which specifies Bradley-Terry strengths $\alpha x^i$:

```{r defineBT}
BT <- function(n,x){
    out <- x^(0:(n-1))
    if(x != 1){
        out <- out*(1-x)/(1-x^n)
    } else {
        out <- out/n
    }
    names(out) <- str_pad(1:n, width=ceiling(log10(n)), pad = "0")
    out
}
BT(4,0.5)
sum(BT(4,0.5))
```

We will consider $n=9$, $x=0.5$, $r=4$ and make repeated observations:


```{r showtabulate}
makeracetable <- function(no_of_races, no_of_competitors, x){
    t(replicate(no_of_races,
                as.numeric(rrace(BT(no_of_competitors, x)))))
}

makeracetable(no_of_races=10, no_of_competitors=5, x=0.7)
```


```{r, label=definegetonecompetitor}
get_one_competitor <- function(racetable, competitor){
    jj <- apply(racetable, 1, function(x){which(x==competitor)})
    tabulate(jj, ncol(racetable))
}
```


```{r, label=exampleracetable, cache=TRUE}
X <- makeracetable(1e4,11,0.7)
get_one_competitor(X,4)
get_one_competitor(X,11)

rbind(
get_one_competitor(X,1) ,
get_one_competitor(X,2) ,
get_one_competitor(X,9) ,
get_one_competitor(X,10)
)

t(sapply(seq_len(11),function(i){get_one_competitor(X,i)}))
```


OK now to put above code in to functional form:

```{r funcformdef}
get_all_competitors <- function(no_of_races, no_of_competitors, x){
    X <- makeracetable(no_of_races=no_of_races, no_of_competitors=no_of_competitors, x=x)
    out <- t(sapply(seq_len(no_of_competitors),function(i){get_one_competitor(X,i)}))
    dimnames(out) <- list(
        true_rank = paste("r_", seq_len(no_of_competitors), sep=""),
        observed_rank = paste("c_", seq_len(no_of_competitors), sep=""))
    return(out/no_of_races)
}
```

Now use it:

```{r useit, cache=TRUE}
get_all_competitors(100,6,0.5)
```

Above, we see from the first line that, with a sample of 100 races,
the top competitor comes first with (estimated) probability about
0.55, second with probability abot 0.25, and so on, coming fifth with
probability 0.01

```{r useit09, cache=TRUE}
get_all_competitors(100, 6, 0.99)
```

Above, all competitors are about the same.  We see a more uniform
spread of observations.

```{r useit01, cache=TRUE}
get_all_competitors(100, 6, 0.1)
```

Above, each competitor is ten times as good as the next one.  We see
underdispersed results.  So what we need to do now is a "meaningful
grid".  This is somewhat computationally intensive.


```{r definegrid}
no_of_races <- 1000
no_of_competitors <- 9

true_x <- seq(from=0.1, by=0.1, to=0.9)
```

First, we combine a whole bunch of X matrices together:

```{r defineallX}
allX <- function(no_of_competitors, no_of_races, true_x){
    out <- NULL
    for(i in seq_along(true_x)){
        out <- rbind(out,
                     cbind(
                         true_x[i],
                         seq_len(no_of_competitors),
                         get_all_competitors(
			 no_of_races = no_of_races,
			 no_of_competitors = no_of_competitors,
			 true_x[i])
                     )
                     )
    }
    rownames(out) <- NULL
    jj <- colnames(out)
    jj[1:2] <- c("x","r")
    colnames(out) <- jj
    return(out)
}
```

```{r makeallX9, cache=TRUE}
allX9 <- allX(
          no_of_competitors = 9,
          no_of_races = 50000,
          true_x = seq(from=0.1, to=0.9, by=0.1)
	  )
```

```{r}
head(allX9)
allX9[seq(from=round(nrow(allX9)/2 - 5), to=round(nrow(allX9)/2 + 5)),]
tail(allX9)
```

Now some observations:


```{r defmakeobs}
make_obs <- function(no_of_observations, no_of_competitors, r_true, x_true){
    kk <- makeracetable(no_of_races = no_of_observations, no_of_competitors = no_of_competitors, x = x_true)
    obs <- tabulate(apply(kk,1,function(o){which(o == r_true)}), nbins=no_of_competitors)
    return(obs)
}
```

```{r tableev, cache=TRUE}
make_obs(no_of_observations=20, no_of_competitors=9, x_true = 0.7, r_true = 6)
make_obs(no_of_observations=20, no_of_competitors=9, x_true = 0.7, r_true = 9)
make_obs(no_of_observations=20, no_of_competitors=9, x_true = 0.1, r_true = 6)


makemaxliketable <- function(n, nobs, no_of_competitors, allX, r_true, x_true){

    maxlike <- function(obs){
        support <- function(i){dmultinom(obs, prob=allX[i, -(1:2)], log=TRUE)}
        p <- sapply(seq_len(nrow(allX)), support)
        allX[which.max(p), 1:2]
    }

    evaluate <- matrix(NA,n,2)
    for(i in seq_len(n)){
        obs <- make_obs(no_of_observations=nobs, no_of_competitors=no_of_competitors, x_true=x_true, r_true=r_true)
        evaluate[i,] <- maxlike(obs)
    }
    return(table(x=evaluate[,1], r=evaluate[,2]))
}
```


```{r label=definemean}
biasx <- function(liketable, x_true){
    values <- as.numeric(rownames(liketable))
    probs <- rowSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - x_true))
}

biasr <- function(liketable, r_true){
    values <- as.numeric(colnames(liketable))
    probs <- colSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - r_true))

}

msex <- function(liketable, x_true){
    values <- as.numeric(rownames(liketable))
    probs <- rowSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - x_true)^2)
}

mser <- function(liketable, r_true){
    values <- as.numeric(colnames(liketable))
    probs <- colSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - r_true)^2)
}
```

```{r label=usebias, cache=TRUE}
(jj <- makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.8))
biasx(jj, x_true = 0.8)
biasr(jj, r_true = 4)
msex(jj, x_true = 0.8)
mser(jj, r_true = 4)
```

### Vary nothing


Just do the same thing a few times to gauge the variability in the system:

```{r varynothing, cache=TRUE}
jj1 <- makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
jj2 <- makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
jj3 <- makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
jj4 <- makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
jj5 <- makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
jj1
jj2
rbind(
c(biasr(jj1, 4), biasx(jj1, 0.7)),
c(biasr(jj2, 4), biasx(jj2, 0.7)),
c(biasr(jj3, 4), biasx(jj3, 0.7)),
c(biasr(jj4, 4), biasx(jj4, 0.7)),
c(biasr(jj5, 4), biasx(jj5, 0.7))
)
```

### Vary `x_true`

```{r execuA, cache=TRUE}
vxt <- list(
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.1),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.2),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.3),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.4),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.5),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.6),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.8),
makemaxliketable(n=4000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.9))
```


```{r}
jj <- rbind(
c(biasr(vxt[[1]], 4), biasx(vxt[[1]], 0.1), mser(vxt[[1]], 4), msex(vxt[[1]], 0.1)),
c(biasr(vxt[[2]], 4), biasx(vxt[[2]], 0.2), mser(vxt[[2]], 4), msex(vxt[[2]], 0.2)),
c(biasr(vxt[[3]], 4), biasx(vxt[[3]], 0.3), mser(vxt[[3]], 4), msex(vxt[[3]], 0.3)),
c(biasr(vxt[[4]], 4), biasx(vxt[[4]], 0.4), mser(vxt[[4]], 4), msex(vxt[[4]], 0.4)),
c(biasr(vxt[[5]], 4), biasx(vxt[[5]], 0.5), mser(vxt[[5]], 4), msex(vxt[[5]], 0.5)),
c(biasr(vxt[[6]], 4), biasx(vxt[[6]], 0.6), mser(vxt[[6]], 4), msex(vxt[[6]], 0.6)),
c(biasr(vxt[[7]], 4), biasx(vxt[[7]], 0.7), mser(vxt[[7]], 4), msex(vxt[[7]], 0.7)),
c(biasr(vxt[[8]], 4), biasx(vxt[[8]], 0.8), mser(vxt[[8]], 4), msex(vxt[[8]], 0.8)),
c(biasr(vxt[[9]], 4), biasx(vxt[[9]], 0.9), mser(vxt[[9]], 4), msex(vxt[[9]], 0.9))
)
colnames(jj) <- c("bias(r)","bias(x)", "mse(r)", "mse(x)")
rownames(jj) <- seq(from=0.1,to=0.9,by=0.1)
jj
```

Above we see  that varying `x_true` from 0.2 to 0.9 increases the
variance of $r_\text{estimated}$

### Vary `r_true`

```{r execuB, cache=TRUE}
makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 1, x_true = 0.7)
makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 3, x_true = 0.7)
makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 5, x_true = 0.7)
makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 9, x_true = 0.7)
```

### Vary `nobs`


```{r execuC, cache=TRUE}
makemaxliketable(n=1000, nobs=10, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
makemaxliketable(n=1000, nobs=30, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
makemaxliketable(n=1000, nobs=80, no_of_competitors=9, allX=allX9, r_true = 4, x_true = 0.7)
```

Above we see that as `nobs` increases, the variance decreases.
