---
title: "exponential Bradley-Terry: likelihood-based methods on a grid of values"
author: "Robin Hankin"
date: "2025-04-02"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("stringr")
library("partitions")
library("magrittr")
```

<font size="1">
(takes maybe twenty minutes to run without cache with $N=9$)
</font>

First we define `N`, the number of competitors.  This is used
consistently in the whole Rmd file, except for the first few bits
where I hard-code five or 11 competitors for ease of exposition.  In
principle, the Rmd file should process whatever the value of `N`.

```{r defineN}
N <- 9  # number of competitors
```


Now we define `BT()` which specifies Bradley-Terry strengths $\alpha x^i$:

```{r defineBT}
BT <- function(n, x){
    out <- x^(0:(n-1))
    if(x != 1){
        out <- out*(1-x)/(1-x^n)
    } else {
        out <- out/n
    }
    names(out) <- str_pad(1:n, width=ceiling(log10(n)), pad = "0")
    out
}
BT(4,0.5)
sum(BT(4,0.5))
```

We will consider $N=9$, $x=0.5$, $r=4$ and make repeated observations:


```{r showtabulate}
makeracetable <- function(no_of_races, no_of_competitors, x){
    t(replicate(no_of_races,
                as.numeric(rrace(BT(no_of_competitors, x)))))
}

makeracetable(no_of_races=10, no_of_competitors=5, x=0.7)
```


```{r, label=definegetonecompetitor}
get_one_competitor <- function(racetable, competitor){
    jj <- apply(racetable, 1, function(x){which(x==competitor)})
    tabulate(jj, ncol(racetable))
}
```


```{r, label=exampleracetable, cache=TRUE}
X <- makeracetable(1e4,11,0.7)  # using N=11 here
get_one_competitor(X,4)
get_one_competitor(X,11)

rbind(
get_one_competitor(X,1) ,
get_one_competitor(X,2) ,
get_one_competitor(X,9) ,
get_one_competitor(X,10)
)

t(sapply(seq_len(11),function(i){get_one_competitor(X,i)}))
```


OK now to put above code in to functional form:

```{r funcformdef}
get_all_competitors <- function(no_of_races, no_of_competitors, x){
    X <- makeracetable(no_of_races=no_of_races, no_of_competitors=no_of_competitors, x=x)
    out <- t(sapply(seq_len(no_of_competitors),function(i){get_one_competitor(X,i)}))
    dimnames(out) <- list(
        true_rank = paste("r_", seq_len(no_of_competitors), sep=""),
        observed_rank = paste("c_", seq_len(no_of_competitors), sep=""))
    return(out/no_of_races)
}
```

Now use it:

```{r useit, cache=TRUE}
get_all_competitors(100, N, 0.5)
```

Above, we see from the first line that, with a sample of 100 races,
the top competitor comes first with (estimated) probability about
0.XXXX, second with probability abot 0.XXX, and so on, coming fifth with
probability 0.XXXXX

```{r useit09, cache=TRUE}
get_all_competitors(1000, N, 0.99)
```

Above, wiht $r=0.99$, all competitors have approximately the same true
BT strength [viz $1,0.99,0.99^2,\ldots, 0.99^9\simeq 0.913$].  We see
a more uniform spread of observations.  Looking at the top row we see
that competitor 1 came first (`c1`) and ninth (`c9`) with about the
same probability.  Now we can use $r=0.1$:

```{r useit01, cache=TRUE}
get_all_competitors(100, N, 0.1)
```

Above, each competitor is ten times as good as the next one.  We see
underdispersed results: competitor $i$ comes $i^\mathrm{th}$ with a
high probability.

What we need to do now is a "meaningful grid".  This is somewhat
computationally intensive.


```{r definegrid}
no_of_races <- 1000
no_of_competitors <- N

true_x <- seq(from=0.1, by=0.1, to=0.9)
```

First, we combine a whole bunch of X matrices together:

```{r defineallX}
allX <- function(no_of_competitors, no_of_races, true_x){
    out <- NULL
    for(i in seq_along(true_x)){
        out <- rbind(out,
                     cbind(
                         true_x[i],
                         seq_len(no_of_competitors),
                         get_all_competitors(
			 no_of_races = no_of_races,
			 no_of_competitors = no_of_competitors,
			 true_x[i])
                     )
                     )
    }
    rownames(out) <- NULL
    jj <- colnames(out)
    jj[1:2] <- c("x","r")
    colnames(out) <- jj
    return(out)
}
```

```{r makeallX9, cache=TRUE}
allX9 <- allX(
          no_of_competitors = N,
          no_of_races = 50000,
          true_x = seq(from=0.1, to=0.9, by=0.1)
	  )
```

```{r}
head(allX9)
allX9[seq(from=round(nrow(allX9)/2 - 5), to=round(nrow(allX9)/2 + 5)),]
tail(allX9)
```

Now some observations:


```{r defmakeobs}
make_obs <- function(no_of_observations, no_of_competitors, r_true, x_true){
    kk <- makeracetable(no_of_races = no_of_observations, no_of_competitors = no_of_competitors, x = x_true)
    obs <- tabulate(apply(kk,1,function(o){which(o == r_true)}), nbins=no_of_competitors)
    return(obs)
}
```

```{r use_make_obs, cache=TRUE}
make_obs(no_of_observations=20, no_of_competitors=N, x_true = 0.7, r_true = 6)
make_obs(no_of_observations=20, no_of_competitors=N, x_true = 0.7, r_true = 9)
make_obs(no_of_observations=20, no_of_competitors=N, x_true = 0.1, r_true = 6)
```

Above, we see vectors of length 9 (because we have $N=9$ competitors,
argument `no_of_competitors`) with elements adding to 20 (because we
have 20 races, argument `no_of_observations`).  The next step is to
extract from such vectors the bias and mean square error)


```{r tableev, cache=TRUE}
makemaxliketable <- function(n, nobs, no_of_competitors, allX, r_true, x_true){

    maxlike <- function(obs){
        support <- function(i){dmultinom(obs, prob=allX[i, -(1:2)], log=TRUE)}
        p <- sapply(seq_len(nrow(allX)), support)
        allX[which.max(p), 1:2]
    }

    evaluate <- matrix(NA,n,2)
    for(i in seq_len(n)){
        obs <- make_obs(no_of_observations=nobs, no_of_competitors=no_of_competitors, x_true=x_true, r_true=r_true)
        evaluate[i,] <- maxlike(obs)
    }
    return(table(x=evaluate[,1], r=evaluate[,2]))
}
```


```{r label=definemean}
biasx <- function(liketable, x_true){
    values <- as.numeric(rownames(liketable))
    probs <- rowSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - x_true))
}

biasr <- function(liketable, r_true){
    values <- as.numeric(colnames(liketable))
    probs <- colSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - r_true))

}

msex <- function(liketable, x_true){
    values <- as.numeric(rownames(liketable))
    probs <- rowSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - x_true)^2)
}

mser <- function(liketable, r_true){
    values <- as.numeric(colnames(liketable))
    probs <- colSums(liketable) / sum(liketable) # sum(probs) == 1
    sum(probs*(values - r_true)^2)
}
```

Now we will illustrate `makemaxliketable()` with some sensible values:

```{r label=usebias1000, cache=TRUE}
(jj <- makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8))
c(
bx = biasx(jj, x_true = 0.8),
br = biasr(jj, r_true = 4),
Mx = msex(jj, x_true = 0.8),
Mr = mser(jj, r_true = 4)
)
```

```{r label=usebias1000a, cache=TRUE}
jj <- makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8)
c(
bx = biasx(jj, x_true = 0.8),
br = biasr(jj, r_true = 4),
Mx = msex(jj, x_true = 0.8),
Mr = mser(jj, r_true = 4)
)
```

```{r label=usebias1000b, cache=TRUE}
jj <- makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8)
c(
bx = biasx(jj, x_true = 0.8),
br = biasr(jj, r_true = 4),
Mx = msex(jj, x_true = 0.8),
Mr = mser(jj, r_true = 4)
)
```

Now try $n=10000$ trials, hopefully a bit more stable:

```{r label=usebias, cache=TRUE}
jj <- makemaxliketable(n=10000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8)
c(
bx = biasx(jj, x_true = 0.8),
br = biasr(jj, r_true = 4),
Mx = msex(jj, x_true = 0.8),
Mr = mser(jj, r_true = 4)
)
```


```{r label=usebias_repeated, cache=TRUE}
jj <- makemaxliketable(n=10000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8)
c(
bx = biasx(jj, x_true = 0.8),
br = biasr(jj, r_true = 4),
Mx = msex(jj, x_true = 0.8),
Mr = mser(jj, r_true = 4)
)
```

```{r label=usebias_repeated2, cache=TRUE}
jj <- makemaxliketable(n=10000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8)
c(
bx = biasx(jj, x_true = 0.8),
br = biasr(jj, r_true = 4),
Mx = msex(jj, x_true = 0.8),
Mr = mser(jj, r_true = 4)
)
```


Above we see a small negative bias in $x$ of about -0.053,
significantly different from zero], effectively zero bias in $r$, a
relatively small MSE for $x$ of about 0.021 and a larger MSE for $r$
of about 2.97.  In this case we have $N=9$ competitors, make 10
observations [that is, have the 9 competitors run 10 races]; and
repeat this many times (here $n=10000$) to stabilize the estimates.

We can try upping the observations to 100 and see if that reduces the
estimation errors:

```{r label=usebias100, cache=TRUE}
(jj <- makemaxliketable(n=10000, nobs=100, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8))
c(
bx = biasx(jj, x_true = 0.8),
br = biasr(jj, r_true = 4),
Mx = msex(jj, x_true = 0.8),
Mr = mser(jj, r_true = 4)
)
```




### Vary nothing

Now I will systemize the above; and again just do the same thing a few
times to gauge the variability in the system:


```{r varynothingTH, cache=TRUE}
jj1 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj2 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj3 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj4 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj5 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj6 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj7 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj8 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj9 <- makemaxliketable(n=30000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
jj1

```{r varynothingbiasmseTH, cache=FALSE}
fish <- rbind(
c(biasr(jj1, 4), biasx(jj1, 0.7), mser(jj1, 4), msex(jj1, 0.7)),
c(biasr(jj2, 4), biasx(jj2, 0.7), mser(jj2, 4), msex(jj2, 0.7)),
c(biasr(jj3, 4), biasx(jj3, 0.7), mser(jj3, 4), msex(jj3, 0.7)),
c(biasr(jj4, 4), biasx(jj4, 0.7), mser(jj4, 4), msex(jj4, 0.7)),
c(biasr(jj5, 4), biasx(jj5, 0.7), mser(jj5, 4), msex(jj5, 0.7)),
c(biasr(jj6, 4), biasx(jj6, 0.7), mser(jj6, 4), msex(jj6, 0.7)),
c(biasr(jj7, 4), biasx(jj7, 0.7), mser(jj7, 4), msex(jj7, 0.7)),
c(biasr(jj8, 4), biasx(jj8, 0.7), mser(jj8, 4), msex(jj8, 0.7)),
c(biasr(jj9, 4), biasx(jj9, 0.7), mser(jj9, 4), msex(jj9, 0.7))
)
colnames(fish) <- c("bias(r)", "bias(x)", "mse(r)", "mse(x)")
```

```{r}
fish
c(mean(fish[,1]), sd(fish[,1])/sqrt(nrow(fish)))
c(mean(fish[,2]), sd(fish[,2])/sqrt(nrow(fish)))
c(mean(fish[,3]), sd(fish[,3])/sqrt(nrow(fish)))
c(mean(fish[,4]), sd(fish[,4])/sqrt(nrow(fish)))
```

Above we have that the bias of $r$ has a standard deviation, given
(approximately at least) by the standard error of the mean.



### Vary `x_true`

```{r execuA, cache=TRUE}
vxt <- list(
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.1),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.2),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.3),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.4),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.5),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.6),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.8),
makemaxliketable(n=40000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.9))
```


```{r}
jj <- rbind(
c(biasr(vxt[[1]], 4), biasx(vxt[[1]], 0.1), mser(vxt[[1]], 4), msex(vxt[[1]], 0.1)),
c(biasr(vxt[[2]], 4), biasx(vxt[[2]], 0.2), mser(vxt[[2]], 4), msex(vxt[[2]], 0.2)),
c(biasr(vxt[[3]], 4), biasx(vxt[[3]], 0.3), mser(vxt[[3]], 4), msex(vxt[[3]], 0.3)),
c(biasr(vxt[[4]], 4), biasx(vxt[[4]], 0.4), mser(vxt[[4]], 4), msex(vxt[[4]], 0.4)),
c(biasr(vxt[[5]], 4), biasx(vxt[[5]], 0.5), mser(vxt[[5]], 4), msex(vxt[[5]], 0.5)),
c(biasr(vxt[[6]], 4), biasx(vxt[[6]], 0.6), mser(vxt[[6]], 4), msex(vxt[[6]], 0.6)),
c(biasr(vxt[[7]], 4), biasx(vxt[[7]], 0.7), mser(vxt[[7]], 4), msex(vxt[[7]], 0.7)),
c(biasr(vxt[[8]], 4), biasx(vxt[[8]], 0.8), mser(vxt[[8]], 4), msex(vxt[[8]], 0.8)),
c(biasr(vxt[[9]], 4), biasx(vxt[[9]], 0.9), mser(vxt[[9]], 4), msex(vxt[[9]], 0.9))
)
colnames(jj) <- c("bias(r)","bias(x)", "mse(r)", "mse(x)")
rownames(jj) <- seq(from=0.1,to=0.9,by=0.1)
jj
matplot(jj,type="b")
cbind(`bias(r)`=jj[,1], `SE(bias)` = jj[,3]/sqrt(4000))
cbind(`bias(x)`=jj[,2], `SE(bias)` = jj[,4]/sqrt(4000))
```


Above we see  that varying `x_true` from 0.1 to 0.9 increases the
variance of $r_\text{estimated}$

### Vary `r_true`

```{r execuB, cache=TRUE}
VR <- list(
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 1, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 2, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 3, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 5, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 6, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 7, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 8, x_true = 0.7),
makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9, r_true = 9, x_true = 0.7)
)
```

```{r showrvary, cache=TRUE}
jj <- rbind(
c(biasr(VR[[1]], 1), biasx(VR[[1]], 0.7), mser(VR[[1]], 1), msex(VR[[1]], 0.7)),
c(biasr(VR[[2]], 2), biasx(VR[[2]], 0.7), mser(VR[[2]], 2), msex(VR[[2]], 0.7)),
c(biasr(VR[[3]], 3), biasx(VR[[3]], 0.7), mser(VR[[3]], 3), msex(VR[[3]], 0.7)),
c(biasr(VR[[4]], 4), biasx(VR[[4]], 0.7), mser(VR[[4]], 4), msex(VR[[4]], 0.7)),
c(biasr(VR[[5]], 5), biasx(VR[[5]], 0.7), mser(VR[[5]], 5), msex(VR[[5]], 0.7)),
c(biasr(VR[[6]], 6), biasx(VR[[6]], 0.7), mser(VR[[6]], 6), msex(VR[[6]], 0.7)),
c(biasr(VR[[7]], 7), biasx(VR[[7]], 0.7), mser(VR[[7]], 7), msex(VR[[7]], 0.7)),
c(biasr(VR[[8]], 8), biasx(VR[[8]], 0.7), mser(VR[[8]], 8), msex(VR[[8]], 0.7)),
c(biasr(VR[[9]], 9), biasx(VR[[9]], 0.7), mser(VR[[9]], 9), msex(VR[[9]], 0.7))
)
colnames(jj) <- c("bias(r)","bias(x)", "mse(r)", "mse(x)")
rownames(jj) <- seq(from=0.1,  to=0.9, by=0.1)
jj
matplot(jj,type="b")
cbind(`bias(r)`=jj[,1], `SE(bias)` = jj[,3]/sqrt(1000))
cbind(`bias(x)`=jj[,2], `SE(bias)` = jj[,4]/sqrt(1000))
```

### Vary `nobs`


```{r execuC, cache=TRUE}
VN <- list(
makemaxliketable(n=1000, nobs= 10, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7),
makemaxliketable(n=1000, nobs= 20, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7),
makemaxliketable(n=1000, nobs= 50, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7),
makemaxliketable(n=1000, nobs=100, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7),
makemaxliketable(n=1000, nobs=200, no_of_competitors=N, allX=allX9, r_true = 4, x_true = 0.7)
)
```


```{r execvarynobs, cache=TRUE}
jj <- rbind(
c(biasr(VN[[1]], 4), biasx(VN[[1]], 0.7), mser(VN[[1]], 4), msex(VN[[1]], 0.7)),
c(biasr(VN[[2]], 4), biasx(VN[[2]], 0.7), mser(VN[[2]], 4), msex(VN[[2]], 0.7)),
c(biasr(VN[[3]], 4), biasx(VN[[3]], 0.7), mser(VN[[3]], 4), msex(VN[[3]], 0.7)),
c(biasr(VN[[4]], 4), biasx(VN[[4]], 0.7), mser(VN[[4]], 4), msex(VN[[4]], 0.7)),
c(biasr(VN[[5]], 4), biasx(VN[[5]], 0.7), mser(VN[[5]], 4), msex(VN[[5]], 0.7))
)
colnames(jj) <- c("bias(r)","bias(x)", "mse(r)", "mse(x)")
rownames(jj) <- c(10,20,50,100,200)
jj
matplot(jj,type="b")
cbind(`bias(r)`=jj[,1], `SE(bias)` = jj[,3]/sqrt(1000))
cbind(`bias(x)`=jj[,2], `SE(bias)` = jj[,4]/sqrt(1000))
```



# A  meaningful grid.

Will keep $N$ fixed at 9 first.  And vary nobs. 


```{r make_meaningful_grid, cache=TRUE}
r_true_values <- 1:9
x_true_values <- seq(from=0.1,to=0.9,by=0.1)
design_matrix <- as.matrix(expand.grid(r_true=r_true_values, x_true = x_true_values))
out <- matrix(-999,nrow(design_matrix),6)
for(i in seq_len(nrow(design_matrix))){
    r_true <- design_matrix[i,1]
    x_true <- design_matrix[i,2]
    jj <-
        makemaxliketable(n=1000, nobs=10, no_of_competitors=N, allX=allX9,
                         r_true = r_true,
                         x_true = x_true
                         )
    out[i,] <- 
        c(
            r_true, x_true,
            biasr(jj, r_true), biasx(jj, x_true), mser(jj, r_true), msex(jj, x_true)
        )
    }
colnames(out) <- c("r_true","x_true", "bias_r", "bias_x", "MSE_r", "MSE_x")
```

```{r show_meaningful_grid}
head(out)
tail(out)
makemat <- function(i){
    m <- matrix(out[,i],length(r_true_values),length(x_true_values))
    rownames(m) <- r_true_values
    colnames(m) <- x_true_values
    return(m)
}
br <- makemat(3)
bx <- makemat(4)
mr <- makemat(5)
mx <- makemat(6)
filled.contour(r_true_values, x_true_values, br, xlab="r", ylab="x", main="Bias(r)")
filled.contour(r_true_values, x_true_values, bx, xlab="r", ylab="x", main="Bias(x)")
filled.contour(r_true_values, x_true_values, mr, xlab="r", ylab="x", main="MSE(r)" )
filled.contour(r_true_values, x_true_values, mx, xlab="r", ylab="x", main="MSE(x)" )

f <- function(...){
    filled.contour(r_true_values, x_true_values, br,
                   xlab="r", ylab="x",
                   main="SHOULD BE FOUR SUBPLOTS")
}
f()

pdf(file="bias_MSE_grid.pdf")
f()
dev.off()
```


```{r foursubplots}
library("cowplot")
f <- function(...){
cowplot::plot_grid(
~filled.contour(r_true_values/N, x_true_values, br,
               xlab = "s", ylab = "x",
               main = "bias(r)"),

~filled.contour(r_true_values/N, x_true_values, bx,
               xlab = "s", ylab = "x",
               main = "bias(x)"),

~filled.contour(r_true_values/N, x_true_values, mr,
               xlab = "s", ylab = "x",
               main = "MSE(r)"),

~filled.contour(r_true_values/N, x_true_values, mx,
               xlab = "s", ylab = "x",
               main = "MSE(x)")
)
}
f()
pdf(file = "bias_MSE_xr.pdf",width=12,height=12)
f()
dev.off()

```
