---
title: "exponential Bradley-Terry: likelihood-based methods on a grid of values"
author: "robin hankin"
date: "2025-04-02"
output: html_document

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(hyper2)
library(stringr)
library("partitions")
library("magrittr")
```

<font size="1">
(takes maybe two or three hours to run without cache; each
<tt>parkrunlikefunobs</tt> chunk takes maybe thirty minutes).
</font>

First we define `BT()` which specifies Bradley-Terry strengths $\alpha x^i$:

```{r defineBT}
BT <- function(n,x){
    out <- x^(0:(n-1))
    if(x != 1){
        out <- out*(1-x)/(1-x^n)
    } else {
        out <- out/n
    }
    names(out) <- str_pad(1:n, width=ceiling(log10(n)), pad = "0")
    out
}
BT(4,0.5)
sum(BT(4,0.5))
```

We will consider $n=9$, $x=0.5$, $r=4$ and make repeated observations:


```{r showtabulate}
makeracetable <- function(no_of_races, no_of_competitors, x){
    t(replicate(no_of_races,
                as.numeric(rrace(BT(no_of_competitors, x)))))
}

makeracetable(10,5,0.7)


```


```{r, label=definegetonecompetitor}
get_one_competitor <- function(racetable, competitor){
    jj <- apply(racetable, 1, function(x){which(x==competitor)})
    tabulate(jj, ncol(racetable))
}
```


```{r, label=exampleracetable, cache=TRUE}
X <- makeracetable(1e4,11,0.7)
get_one_competitor(X,4)
get_one_competitor(X,11)

rbind(
get_one_competitor(X,1) ,
get_one_competitor(X,2) ,
get_one_competitor(X,9) ,
get_one_competitor(X,10)
)

t(sapply(seq_len(11),function(i){get_one_competitor(X,i)}))
```


OK now to put above code in to functional form:

```{r funcformdef}
get_all_competitors <- function(no_of_races, no_of_competitors, x){
    X <- makeracetable(no_of_races, no_of_competitors, x)
    out <- t(sapply(seq_len(no_of_competitors),function(i){get_one_competitor(X,i)}))
    dimnames(out) <- list(
        true_rank = paste("r_", seq_len(no_of_competitors), sep=""),
        observed_rank = paste("c_", seq_len(no_of_competitors), sep=""))
    return(out/no_of_races)
}
```

Now use it:

```{r useit, cache=TRUE}
get_all_competitors(100,6,0.5)
```

Above, we see from the first line that, out of 100 races, the top
competitor comes first with (estimated) probability 0.53 second with
probability 0.25 times, and so on, coming fifth with probability 0.01

```{r useit09, cache=TRUE}
get_all_competitors(100,6,0.99)
```

Above, all competitors are about the same.  We see a more uniform
spread of observations.


```{r useit01, cache=TRUE}
get_all_competitors(100, 6, 0.1)
```

Above, each competitor is ten times as good as the next one.   We see
underdispersed results.

So what we need to do now is a "meaningful grid".  This is somewhat
computationally intensive.


```{r definegrid}
no_of_races <- 1000
no_of_competitors <- 9

true_x <- seq(from=0.1, by=0.1, to=0.9)
true_rank <- c(1, 3, 5, 7, 9)
a <- as.matrix(expand.grid(true_x, true_rank))
```

First, we combine a whole bunch of X matrices together:

```{r defineallX}
allX <- function(no_of_competitors, no_of_races, true_x){
    out <- NULL
    for(i in seq_along(true_x)){
        out <- rbind(out,
                     cbind(
                         true_x[i],
                         seq_len(no_of_competitors),
                         get_all_competitors(no_of_races,  no_of_competitors, true_x[i])
                     )
                     )
    }
    rownames(out) <- NULL
    jj <- colnames(out)
    jj[1:2] <- c("r","n")
    colnames(out) <- jj
    return(out)
}
```

```{r makeallXjj, cache=TRUE}
jj <- allX(9, 10000, seq(from=0.1,to=0.9,by=0.1))
    print("bug here")
```

```{r}
head(jj)
tail(jj)
```

Now some observations:


```{r defmakeobs}
make_obs <- function(no_of_observations, r_true, x_true){
    kk <- makeracetable(no_of_observations, r_true, x_true)

    print("bug here")
    obs <- tabulate(apply(kk,1,function(o){which(o == r_true)}),nbins=no_of_competitors)
    return(obs)
}
```

```{r tableev}
make_obs(no_of_observations=20, x_true = 0.7, r_true = 6)
make_obs(no_of_observations=20, x_true = 0.7, r_true = 9)
make_obs(no_of_observations=20, x_true = 0.1, r_true = 6)

print("bug here")
maxlike <- function(obs){
    support <- function(i){dmultinom(obs, prob=jj[i,-(1:2)], log=TRUE)}
    p <- sapply(seq_len(nrow(jj)), support)
    jj[which.max(p), 1:2]
}

n <- 1000
evaluate <- matrix(NA,n,2)
r_true <- 5
x_true <- 0.6
for(i in seq_len(n)){
    obs <- make_obs(no_of_observations=80, r_true, x_true)
print("bug here")
    evaluate[i,] <- maxlike(obs)
}
table(evaluate[,1], evaluate[,2])
```
