---
documentclass: jss
author:
  - name: Robin K. S. Hankin
    affiliation: Auckland University of Technology
    address: >
      2-14 Wakefield Street
      Auckland, New Zealand
    email: \email{hankin.robin@gmail.com}
    url: http://aut.ac.nz/
title:
  formatted: "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  plain:     "An Objective Assessment of Points Scorings in Formula 1 Motor Racing"
  # For running headers, if needed
  short:     "Formula 1 Points Scoring"
abstract: >
  The points scoring system of Formula 1 motor racing is a long-standing
  contention among fans and competitors, with changes to the points
  scored occurring frequently.  Here, I apply generalized Plackett-Luce
  likelihood functions to Formula 1 racing, and assess historical points
  scoring systems objectively.  In a well-defined sense, the points
  scoring system of 1985 is the best of the historical ones.  I make
  similar observations about related competitive situations including
  the Eurovision Song contest and XXX.
keywords:
  # at least one keyword must be supplied
  formatted: [keywords, not capitalized, "\\proglang{Java}"]
  plain:     [keywords, not capitalized, Java]
preamble: >
  \usepackage{amsmath}
output: rticles::jss_article
---

# Introduction

Formula 1 racing is an important and prestigious motor sport
\citep{moul2009}.  Season ranking is based on a points system; points
are awarded to drivers and teams based on race order, and accumulate
additively.  The current points system, for example, is
(25-18-15-12-10-8-6-4-2-1) [that is, the winner is awarded 25 points,
second place 18 points, and so on], although since 2019 the fastest
lap attracts one point.

Given that racing is a zero-sum game---and that points are
monotonically decreasing---each player will try to get as high a rank
as possible regardless of the actual points system used.  However,
there are other consistent interpretations.  \cite{bakhrankova2011},
for example, considers the possibility of inter-driver collusion, a
phenomenon not pursued here; and authors such as
\citeauthor{mastromarco2009} suggest that the frequency of rule
changes is driven by factors such as driver safety and revenue
optimization.

Points systems similar to that of Formula 1 are common in other sports
and examples would include modern pentathlon, curling, chess, rowing,
and figure skating; all have the common feature of interpretation of
an order statistic to a points.  Further, we see points systems
frequently in the wider context of competitive situations such as the
Eurovision Song Contest, Australian MasterChef, University rankings,
educational league tables, and car safety ratings.

## Bradley-Terry and generalizations for rank statistics

The Bradley-Terry model \citep{bradley1952} assigns non-negative
strengths $p_1,\ldots, p_n$ to each of $n$ competitors in such a way
that the probability of $i$ beating $j\neq i$ in pairwise competition
is $\frac{p_i}{p_i+p_j}$; it is conventional to normalize so that
$\sum p_i=1$.  Further, we use a generalization due to
\citet{luce1959}, in which the probability of competitor $i$ winning
in a field of $j=1,\ldots, n$ is $\frac{p_i}{p_1+\cdots +p_n}$.
Noting that there is information in the whole of the finishing order,
not just the first across the line, we can follow \cite{plackett1975}
and consider the runner-up to be the winner among the remaining
competitors; and so on down the finishing order.  Without loss of
generality, if the order of finishing were $1,2,3,4$, then a suitable
\citeauthor{plackett1975} likelihood function would be

\begin{equation}\label{competitors_1_to_4_likelihood}
\frac{p_1}{p_1+p_2+p_3+p_4}\cdot
\frac{p_2}{p_2+p_3+p_4}\cdot
\frac{p_3}{p_3+p_4}\cdot
\frac{p_4}{p_4}\cdot
\end{equation}

We now use a technique due to \cite{hankin2010,hankin2017} and
introduce fictional (reified) entities whose nonzero Bradley-Terry
strength helps certain competitors or sets of competitors.  The
canonical example would be the home-ground advantage in association
football.  If players (teams) $1,2$ with strengths $p_1,p_2$ compete,
and if our observation were $a$ home wins and $b$ away wins for team
$1$, and $c$ home wins and $d$ away wins for team $2$, then a suitable
likelihood function would be

\[
\left(\frac{p_1+p_H}{p_1+p_2+p_H}\right)^a
\left(\frac{p_1}{p_1+p_2+p_H}\right)^b
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^c
\left(\frac{p_2+p_H}{p_1+p_2+p_H}\right)^d
\]

where $p_H$ is a quantification of the beneficial home ground effect.
Similar techniques have been used to account for the first-move
advantage in chess, and effective coordination between members of
doubles tennis teams; we may use a similar device to account for pole
position in Formula 1.  Here I analyse racing results from the 2017
season using the `hyper2` package \citep{hankin2017} which implements
the Plackett-Luce likelihood function with additional reified
entities.

# Formula 1 dataset


Here we use the results from 2017:

\small{
\begin{CodeChunk}
\begin{CodeOutput}
           AUS CHN BHR RUS ESP MON CAN AZE AUT GBR HUN BEL ITA SIN MAL JPN USA MEX BRA ABU
Hamilton     2   1   2   4   1   7   1   5   4   1   4   1   1   1   2   1   1   9   4   2
Vettel       1   2   1   2   2   1   4   4   2   7   1   2   3 Ret   4 Ret   2   4   1   3
Bottas       3   6   3   1 Ret   4   2   2   1   2   3   5   2   3   5   4   5   2   2   1
Raikkonen    4   5   4   3 Ret   2   7  14   5   3   2   4   5 Ret DNS   5   3   3   3   4
Ricciardo  Ret   4   5 Ret   3   3   3   1   3   5 Ret   3   4   2   3   3 Ret Ret   6 Ret
...
Hartley      0   0   0   0   0   0   0   0   0   0   0   0   0   0   0   0  13 Ret Ret  15
Button       0   0   0   0   0 Ret   0   0   0   0   0   0   0   0   0   0   0   0   0   0
Resta        0   0   0   0   0   0   0   0   0   0 Ret   0   0   0   0   0   0   0   0   0
\end{CodeOutput}
\end{CodeChunk}
}

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourth in Russia, and so on (Hartley,
Button, and Resta placed last).  In the first column, we see the
result from Australia (AUS) in which Hamilton came second, Vettel
first, Bottas third, and so on.  Notation used also includes six
different classes of no-score such as `Ret` for retired, `WD` for
withdrawn, and so on; a zero entry means "did not finish".  It is
straightforward to translate this table into a Plackett-Luce
log-likelihood function using the `hyper2` package; for simplicity we
will consider only the 10 top-ranked drivers\footnote{In the
Plackett-Luce likelihood function, the performance of lower-ranked
players can be informative about higher-ranked players' strengths.
For example, we see that Vettel retired twice (in Singapore and
Japan), so any player who placed in those venues will effectively
``steal'' strength from Vettel, and generally ``give'' it to Hamilton
or Bottas.}.  Although it has many terms, the first few are

```{r,echo=FALSE,print=FALSE}
library(hyper2,quietly=TRUE)
H <- ordertable2supp(keep(F1_table_2017,1:10))
head(H,5)
```

Finding the maximum likelihood estimate for the players' strengths is straightforward:

```{r,echo=FALSE}
mH <- maxp(H)
print(mH)
pie(maxp(H)[c(1,6,2,7,3,8,4,9,5,10)])
```

We see the that the top three players, Hamilton, Vettel, Bottas, are
of roughly equal strength and indeed a likelihood ratio test
[`samep.test()`] does not reject the null of equality ($p=0.72$).

## Likelihood scoring vs points scoring

The points scored by each driver give a ranking shown in the table;
but Plackett-Luce likelihoods also give a ranking that us unencumbered
by the arbitrary points given.  We can plot one ranking against the other:

```{r,label=likevspoints,cache=TRUE}
jj <- read.table("formula1_2017.txt",header=TRUE)
points <- jj[,21]
names(points) <- rownames(jj)

mL2017 <- maxp(H)

m <- mL2017   # maxp(H)
ox <- round(rank(-points)[1:10]) # negative because number 1 = largest number of points
oy <- rank(-m)      # negative because number 1 = largest strength
par(pty='s')        # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,10.2),ylim=c(0,10.2),pch=16,
xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE)       # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(ox)[i],pos=4,col='gray') }
par(xpd=FALSE)      # stop diagonal line from protruding beyond plotting region
abline(0,1)
grid()
print(ox)
print(oy)
```

\bibliography{chess}
