---
title: "formula1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

## Formula 1

Formula 1 racing is an important and prestigious motor sport.  Here I
analyse a series of racing results using the `hyper` package which
implements the Plackett-Luce likelihood function.

Consider the following dataset, taken from Wikipedia:

```{r cars}
f2017 <- read.table("formula1_2017.txt",header=TRUE)
head(f2017)
```

Each row is a driver and each column (after the first) a venue.  We
see that Hamilton, the first row, came second in Australia, first in
China, second in Bahrain, fourthin Russia, and so on.  In the first
columnm, we see the result from Australia in which Hamilton came
second, Vettel first, Bottas third, and so on.  Notation used also
includes six different classes of no-score such as "Ret" for retired,
"WD" for withdrawn, and so on.  I now define a bespoke R function that
translates data of this type into a log-likelihood function using the
`hyper2` package.  The function is similar to, but slightly different
from, the analysis in `eurovision.R`.



```{r,defF1}
F1_likelihood <- function(wiki_table){

  noscore <- c("Ret", "WD", "DNS", "DSQ", "DNP", "NC")
  first_and_last <- c(1,ncol(wiki_table))

  racers <- wiki_table[,1]
  venues <- colnames(wiki_table)[-first_and_last] 

  ## Now create a numeric matrix, fmat.  Two steps: first, count
  ## any no-score as zero:
  
  f <- function(x){
    if(any(x %in% noscore)){x[x%in%noscore] <- 0}
    return(x)
  }

  jj <- apply(wiki_table,2,f)
  
  ## Second, convert to numeric and strip out names; transpose of
  ## wiki_table (because we want each row to be a venue):

  fmat <- matrix(as.numeric(jj[,-first_and_last]),byrow=TRUE,ncol=nrow(wiki_table)) 
  colnames(fmat) <- racers
  rownames(fmat) <- venues

  F1 <- hyper2(d=ncol(fmat))

  ## Now cycle through the rows; each row is a venue [voter]
  for(i in seq_len(nrow(fmat))){
    d <- fmat[i,,drop=TRUE]
    while(any(d>0)){
      eligible <- which(d>=0)
      
      ## The first choice among eligible players has +1 power on the
      ## numerator:
      F1[which(d==1)] %<>% "+"(1)

      ## denominator of all eligible players; power -1
      F1[eligible] %<>% "-"(1)

      ## once you've come first in the field, you are ineligible to be first again:
      d[d==1] <- -1
      
      ## everyone moves down the list, so who *was* in second place
      ## becomes first place, who *was* third place becomes second,
      ## and so on:
      d[d>0] %<>% "-"(1)

    } # while() loop closes
  } # i loop closes
  
  ## syntatic sugar:
  pnames(F1) <- racers

  return(F1)
}
```

We can apply this to the dataset:

```{r,applydata}
L2017 <- F1_likelihood(f2017)
head(L2017,3)
```

Above, we see a small part of the Placket-Luce likelihood function for
the 2017 results.  We can find the maximum likelihood estimator by
using the `maxp()` function:


```{r,maxp2017}
dotchart(maxp(L2017),pch=16,main='2017 Formula 1')
dotchart(log(maxp(L2017)[1:23]),pch=16,main='2017 Formula 1, log scale')
```

# Likelihood scoring vs points scoring

The championship winning order is determined by a points system: the
winner is awarded 25 points, second place 18 points, and so on.  We
can now check to see how the points total matches up with the
likelihood estimation.

```{r,likelihoodandscore}
points <- f2017$points
names(points) <- f2017$driver

m <- maxp(L2017)
ox <- order(points,decreasing=TRUE)
oy <- order(m,decreasing=TRUE)
par(pty='s') # square plot
plot(ox,oy,asp=1,pty='s',xlim=c(0,25),ylim=c(0,25),pch=16,
xlab="official order",ylab="my order",main='Formula 1, 2017 season')
par(xpd=TRUE) # allow drivers' names to appear outside plotting region
for(i in seq_along(ox)){text(ox[i],oy[i],names(m)[ox[i]],pos=4,col='gray') }
par(xpd=FALSE) # stop diagonal line from protruding beyond plotting region
abline(0,1)
```

In the graph above, we see that both systems consider Hamilton to be
the best.  However, F1 points place Vettel second and Bottas third,
while `hyper2` places Bottas second and Vettel third.  Overall, there
is a high degree of correlation between the two, but differences in
detail.

# Analysis of nulls:

## Null of equal strengths:

```{r,equalstrength}
L1 <- loglik(indep(equalp(L2017)),L2017)
L2 <- loglik(indep(maxp(L2017)),L2017)
pchisq(2*(L2-L1),df=25-1,lower.tail=FALSE)
```

highly signficant!

## Null of Hamilton being average


This script takes about a minute to run.  It creates hyper2 objects
`F1_2014` through `F1_2017`, which are the likelihood functions for
the F1 results table for each of the years.

Function `F1_likelihood()` is defined here; this creates objects such as
`F1_2017` by processing `formula1_2017.txt`.

At the end of the script, hyper2 object `F1_total` is created, which
is a likelihood function for all four years' data.  This object is
called 'formula1' in the package.  File `F1_allyears.R` is functionally
identical but uses slightly different R idiom.


