---
title: "The ordertrans() function"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("magrittr")
```

# Introduction

The `ordertrans()` function can be difficult to understand and this
short document provides some sensible use-cases.



## Hello world

The manpage provides a very simple example:


```{r,label=simpexample}
x <- c(d=2,a=3,b=1,c=4)
x
```

In the above, object `x` is a named vector with elements
`seq_along(x)` in some order.  Technically `x` is an order vector
because it answers the question "where did the first-named competitor
come?"  (see also the discussion at `rrank.Rd`).  However, it might
equally be a rank vector because it answers the question "who came
first? who came second?".

But note that `x` is not helpfully structured to answer either of
these questions: you have to go searching through the names or the
ranks respectively---both of which may appear in any order---to find
the competitor or rank you are interested in.  What we want is to
rearrange the elements of `x` so that the names are in alphabetical
order.  That way we can see straighaway where competitor `a` placed
(in this case, third).  This nontrivial task is accomplished by
function `ordertrans()`:


```{r,label=useordertrans}
o <- ordertrans(x,letters[1:4])
o
```

Observe that objects `x` and `o` are equal in the sense that they are
a rearrangement of one another:

```{r,label=equalrearrangement}
identical(x, o[c(4,1,2,3)])
identical(o, x[c(2,3,4,1)])
```

One consequence of this is that the resulting Plackett-Luce support
functions will be the same:

```{r,showordervec}
ordervec2supp(x)
ordervec2supp(o)
```

Note carefully that the two support functions above can be
mathematically identical but not formally `identical()` because
neither the order of the brackets, nor the order of terms within a
bracket, is defined.  They look the same on my system but YMMV.  It is
possible that the two support functions might appear to be different
even though they are mathematically the same. We can verify that the
two are mathematically identical using package idiom `==`:

```{r}
ordervec2supp(x) == ordervec2supp(o)
```

## Skating

Let us consider the skating dataset and use `ordertrans()` to study
it.

```{r,label=lookatskating}
skating_table
```

We might ask how judges 1 and 2 compare to one another?  We need to
create vectors like `x` and `y` above:

```{r,makej1j2}
j1 <- skating_table[,1]  # column 1 is judge number 1
names(j1) <- rownames(skating_table)
j2 <- skating_table[,2]  # column 2 is judge number 2
names(j2) <- rownames(skating_table)
j1
j2
cbind(j1,j2)
```

In the above, see how objects `j1` and `j2` have identical names, in
the same order.  Observe that `hughes` is ranked 1 by judge 1, and 4
by judge 2.  This makes it sensible to plot `j1` against `j2`:

```{r}
par(pty='s')  # forces plot to be square
plot(j1,j2,asp=1,pty='s',xlim=c(0,25),ylim=c(0,25),pch=16,xlab='judge 1',ylab='judge 2')
abline(0,1)  # diagonal line
for(i in seq_along(j1)){text(j1[i],j2[i],names(j1)[i],pos=4,col='gray',cex=0.7)}
```

In the above plot we see general agreement but differences in detail.
For example, `hughes` is ranked first by judge 1 and fourth by judge
2.  However, other problems are not so easy.  Suppose we wish to
compare the ranks according to likelihood with the ranks according to
some points system.

```{r,label=maxlikeskating,cache=TRUE}
mL <- maxp(skating)  # 'L' for Likelihood
mL
```

Note that in the above, the competitors' names are in alphabetical
order.  We first need to convert strengths to ranks:

```{r,label=strengthstoranks}
mL[] <- rank(-mL)  # minus because ranks orders from weak to strong
mL
```

(note that the names are in the same order as before, alphabetical).
In the above we see that `slutskya` ranks first, `hughes` second, and
so on.  Another way of ranking the skaters is to use a Borda-type
system: essentially the rowsums of the table (and, of course, the
_lowest_ score wins):


```{r,label=pointsskating,cache=TRUE}
mP <- rowSums(skating_table)  # 'P' for Points
mP[] <- rank(mP,ties='first') # positive sign here
mP
```

It is not at all obvious how to compare `mP` and `mL`.  For example,
we might be interested in `hegel`.  It takes some effort to find that
her likelihood rank is 20 and her Borda rank is 19.  Function
`ordertrans()` facilitates this:


```{r,label=ordertransexample}
ordertrans(mP,names(mL))  
```

See above how `ordertrans()` shows the points-based ranks but in
alphabetical order, to facilitate comparison with `mL`.  We can now
plot these against one another:


```{r,label=crapplot}
plot(mL,ordertrans(mP,mL))
```

However, the plot above is a bit crude.  Function `ordertransplot()`
gives a more visually pleasing output:


```{r,label=showoffordertransplot}
ordertransplot(mL,mP,xlab="likelihood rank",ylab="Borda rank")
```

So now we may compare judge 1 against likelihood:


```{r,transplotjudge1}
ordertransplot(mL,j1,xlab="likelihood rank",ylab="Judge 1 rank")
```


In the above, looking at the lower-left corner, we see (reading
horizontally) that the likelihood method placed Slutskya first, then
Hughes second, then Kwan third; while (reading vertically) judge 1
placed Hughes first, then Kwan, then Slutskya.
