---
title: "plackett_luce_monster"
output: html_document
date: '2022-05-16'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("hyper2")
library("partitions")
```

We will calculate a Plackett-Luce likelihood function for four runners
$\left\lbrace 1,2,3,4\right\rbrace$ but with a reified Bradley-Terry
monster that corresponds to having subsets of mutual friends who
support one another.  For example, we might have runners 1 and 2 being
friends, and runners 3 and 4 being friends: we would represent this as
$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
3,4\right\rbrace\right\rbrace$.

The statistical model is as follows.  If a runner finishes, then he
cheers on all his friends who are still running, thereby adding to
their strength.  This is a very simple form of non-independence.
 
First do three runners, and suppose the equivalence classes are
$\left\lbrace\left\lbrace 1,2\right\rbrace,\left\lbrace
3\right\rbrace\right\rbrace$.


 123: $\frac{p_1}{p_1+p_2+p_3}\cdot\frac{p_2+M}{p_2+p_3+M}\qquad\frac{p_1}{p_1+p_2+p_3}\cdot\frac{\lambda p_2}{\lambda p_2+p_3}$
 
 132: $\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+p_2+M}\qquad\frac{p_1}{p_1+p_3+p_2}\cdot\frac{p_3}{p_3+\lambda p_2}$
 
 213: $\frac{p_2}{p_2+p_1+p_3}\cdot\frac{p_1+M}{p_1+p_3+M}\qquad\frac{p_2}{p_2+p_1+p_3}\cdot\frac{\lambda p_1}{\lambda p_1+p_3}$
 
 231: $\frac{p_2}{p_2+p_3+p_1}\cdot\frac{p_3}{p_3+p_1+M}$
 
 312: $\frac{p_3}{p_3+p_1+p_2}\cdot\frac{p_1}{p_2+p_1}$
 
 321: $\frac{p_3}{p_3+p_2+p_1}\cdot\frac{p_2}{p_2+p_1}$


Now four runners with classes $\left\lbrace\left\lbrace
1,2\right\rbrace,\left\lbrace 3,4\right\rbrace\right\rbrace$.

 1234: $\frac{1}{1+2+3+4}\cdot\frac{2+M}{2+3+4+M}\cdot\frac{3}{3+4}$

 1243: $\frac{1}{1+2+4+3}\cdot\frac{2+M}{2+4+3+M}\cdot\frac{4}{4+3}$

 1324: $\frac{1}{1+3+2+4}\cdot\frac{3}{3+2+4+M}\cdot\frac{2}{2+4+M}$

 1342: $\frac{1}{1+3+4+2}\cdot\frac{3}{3+4+2+M}\cdot\frac{4+M}{2+4+M}$

 1423: $\frac{1}{1+4+2+3}\cdot\frac{4}{4+2+3+M}\cdot\frac{2}{2+3}$

 1432: $\frac{1}{1+4+3+2}\cdot\frac{4}{4+3+2+M}\cdot\frac{3}{2+3}$

 2134: $\frac{2}{2+1+3+4}\cdot\frac{1+M}{1+3+4+M}\cdot\frac{3}{3+4}$

 2143: $\frac{2}{2+1+4+3}\cdot\frac{1+M}{1+4+3+M}\cdot\frac{4}{4+3}$

 2314: $\frac{2}{2+3+1+4}\cdot\frac{3}{3+1+4+M}\cdot\frac{1}{1+4+M}$

 2341: $\frac{2}{2+3+4+1}\cdot\frac{3}{3+4+1+M}\cdot\frac{4+M}{1+4+M}$

 2413: $\frac{2}{2+4+1+3}\cdot\frac{4}{4+1+3+M}\cdot\frac{1}{1+3}$

 2431: $\frac{2}{2+4+3+1}\cdot\frac{4}{4+3+1+M}\cdot\frac{3}{1+3}$

etc.


Now 123 tethered, 45 tethered:


12345: $\frac{1}{1+2+3+4+5}\cdot\frac{2+M}{2+3+4+5+M}\cdot\frac{3+M}{3+4+5+M}\cdot\frac{4}{4+5}$

12354: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{3+M}{3+5+4+M}\cdot\frac{5}{5+4}$

12435: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{3}{3+5+M}$

12453: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+3+5+4+M}\cdot\frac{4}{3+5+4+M}\cdot\frac{5+M}{5+3+M}$

12534: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+3+4+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{3}{3+4+M}$

12543: $\frac{1}{1+2+3+5+4}\cdot\frac{2+M}{2+5+4+3+M}\cdot\frac{5}{3+5+4+M}\cdot\frac{4+M}{4+3+M}$



Might be better to consider ${5\choose 2\,3}$ arrangements of $aaabb\ldots bbaaa$.

aaabb: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{a+M}{a+b+b+M}\cdot\frac{b}{b+b+M}$

aabab: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{a}{a+b+M}$

aabba: $\frac{a}{a+a+a+b+b}\cdot\frac{a+M}{a+a+b+b+M}\cdot\frac{b}{a+b+b+M}\cdot\frac{b+M}{b+a+M}$

abaab: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$
 
ababa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{a}{a+a+b+M}\cdot\frac{b}{b+a+M}$

abbaa: $\frac{a}{a+a+a+b+b}\cdot\frac{b}{a+a+b+b+M}\cdot\frac{b+M}{b+a+a+M}\cdot\frac{a}{a+a+M}$

baaab: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{a+M}{a+b+M}$
 
baaba: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{a+M}{a+a+b+M}\cdot\frac{b}{a+b+M}$
 
babaa: $\frac{b}{a+a+a+b+b}\cdot\frac{a}{a+a+a+b+M}\cdot\frac{b}{b+a+a+M}\cdot\frac{a}{a+a}$
 
bbaaa: $\frac{b}{a+a+a+b+b}\cdot\frac{b+M}{a+a+a+b+M}\cdot\frac{a}{a+a+a}\cdot\frac{a}{a+a}$



```{r,label=definefunctionf}
f <- function (a,b,M){
return(factorial(3)*factorial(2)*(
(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(a+M)/(a+b+b+M)*(b)/(b+b+M)+
(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b)/(a+b+b+M)*(a)/(a+b+M)+
(a)/(a+a+a+b+b)*(a+M)/(a+a+b+b+M)*(b)/(a+b+b+M)*(b+M)/(b+a+M)+
(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(a)/(a+a+b+M)*(a+M)/(a+b+M)+
(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(a)/(a+a+b+M)*(b)/(b+a+M)+
(a)/(a+a+a+b+b)*(b)  /(a+a+b+b+M)*(b+M)/(b+a+a+M)*(a)/(a+a+M)+
(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(a+M)/(a+a+b+M)*(a+M)/(a+b+M)+
(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(a+M)/(a+a+b+M)*(b)/(a+b+M)+
(b)/(a+a+a+b+b)*(a)  /(a+a+a+b+M)*(b)/(b+a+a)*(a)/(a+a)+
(b)/(a+a+a+b+b)*(b+M)/(a+a+a+b+M)*(a)/(a+a+a)*(a)/(a+a)))
}
```


```{r}
f(4.4,3.3, 0.2)*12
f(4.4,3.3, 0.0)*12
```