---
output:
  pdf_document: default
  html_document: default
---
# The importance of serve in table tennis.

Suppose We have three players, "a","b", and "c" who play table tennis.
We seek to quantify the effect of the serve.  The scoreline is as
follows:

* A vs B, A serves, 5-1
* A vs B, B serves, 1-3
* A vs C, A serves, 4-1
* A vs C, C serves, 1-2

(note that A does not play C).  This dataset is easily analysed with
the ```hyper2``` package.

```{r}
library("hyper2",quietly=TRUE)
```

The first step is to define a ```hyper2``` object:

```{r}
H <- hyper2(pnames=c("S","a","b","c"))
```

Now we have to input the data:


### A vs B, A serves, score 5-1:
```{r}
H[c("a","S")]     %<>% "+"(5  ) # a+S win 5 games
H["b"]            %<>% "+"(  1) # b wins 1 game
H[c("a","b","S")] %<>% "-"(5+1) # a+b+S play 1+5=6 games
```

### A vs B, B serves, score 1-3:
```{r}
H[c("b","S")]     %<>% "+"(3  ) # b+S win 3 games
H["a"]            %<>% "+"(  1) # a wins 1 game 
H[c("a","b","S")] %<>% "-"(3+1) # a+b+S play 1+3=4 games
```

### A vs C, A serves, score 4-1:
```{r}
H[c("a","S")]     %<>% "+"(4  ) # a+S win 4 games
H["c"]            %<>% "+"(  1) # c wins 1 game 
H[c("a","c","S")] %<>% "-"(4+1) # a+c+S play 1+4=5 games
```

### A vs C, C serves, score 1-2:
```{r}
H["a"]            %<>% "+"(1  ) # a wins 1 game
H[c("c","S")]     %<>% "+"(  2) # c+S win 2 games
H[c("a","c","S")] %<>% "-"(1+2) # a+c+S play 1+2=3 games
```

So H is the likelihood function for three players and a serve.

# Results 
We now find the unconstrained MLE:

```{r}
p1 <- maxp(H)
p1
```

And we can find the support at this point:

```{r}
maxlike_free <- loglik(H,indep(p1))
maxlike_free
```

Now find the MLE conditional on the serve having zero strength:

```{r}
small <- 1e-6
maxlike_constrained <-
    constrOptim(
        f = function(p){loglik(H,p)},
        theta=c(small, 1/3-small/2, 1/3-small/2),
        grad=NULL,
        ui=rbind(diag(3),-diag(3),-1),
        ci=c(rep(small/2,3),c(-small*2,-1,-1),-1),
        control=list(fnscale = -1)  # finds max, not min
    )
```	

The difference in support is:

```{r}    	    
delta_support <- maxlike_free - maxlike_constrained$value
delta_support
```

and the likelihood ratio is:
```{r}
LR <- exp(delta_support)
LR
```

On the null, Wilks's theorem says that $\Lambda\sim\chi^2_1$, where
$\Lambda=2\log(L_1/L_2)$, so the
p-value is

```{r}
Lambda <- 2*delta_support
pval <- pchisq(Lambda,df=1,lower.tail=FALSE)
pval 
```



The difference in support, ```delta_support```, is about 2.62 which
exceeds Edwards's two-units-of-support criterion.  The likelihood
ratio LR is about 13.7, and Wilks's theorem gives us that
$\Lambda=2\log(LR)$ is distributed as chi-squared with one degree of
freedom; the tail area, pval, is about 0.022 which gives us a
significant result.


Further, note that our maximum likelihood estimate for strengths
allows us to assess what might happen when A plays C (which was not
observed).  For convenience, here is the evaluate:

```{r}
maxp(H)
```

If A serves, we would have $A+S:C\simeq 0.24+0.45:0.16\equiv 69:16$,
or A wins with probability of about $0.81$.  If C serves we have
$A:C+S\simeq 0.24:0.16+0.45\equiv 24:61$, so A wins with a probability
of about $0.28$.