---
title: "ternary_hyper2"
author: "Robin Hankin"
date: "2022-11-21"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Ternary plots for hyper2 samples

Here I show how to use the excellent `Ternary` package to generate
triangular plots for `hyper2` likelihood functions with size 3.  First
load the `Ternary` and `hyper2` packages:

```{r loadlibraries}
library("hyper2")
library("Ternary")
```

## The `chess` dataset

Now generate a synthetic dataset.  Here we will sample from the
posterior of a uniform prior with a likelihood function specified by
the `chess` object (this takes quite a long time):

```{r generatedata, cache=TRUE}
set.seed(0)
xx <- rp(1000,chess)
```

Using the `TernaryPlot()` and `TernaryPoints()` functions is straightforward:
```{r plotdata}
jj <- pnames(chess)
TernaryPlot(atip=jj[1],btip=jj[2],ctip=jj[3])
TernaryPoints(xx,cex=0.5)
TernaryPoints(chess_maxp,pch=16,col="red")
```

Above we see black circles being samples from the posterior and the
red point is the maximum likelihood estimate for the three players'
strengths.

We can adapt this technique to higher-dimensional `hyper2` objects.


## The `icons` dataset

Consider the `icons` likelihood function:

```{r showicons}
icons
```


We can sample from this object and plot it in different ways:


```{r samplefromicons,cache=TRUE}
n <- 10000
x <- rp(n,icons)
x[sample(n,5),]
```


```{r, iconsmatplot}
matplot(x,type="l",lty=1)
legend("topright",col=1:6,lty=1,legend=pnames(icons))
```


```{r, selectjustbigones}
wanted <- c(1,3) # NB,PB
xx <- cbind(x[,wanted],fillup=rowSums(x[,-wanted]))
TernaryPlot(atip="NB",btip="PB",ctip="other")
TernaryPoints(xx,cex=0.5)
```



